---
title: 'Lab 6 - Part 2: Chi-Square w/lizards tail'
author: "Karina Johnston"
date: "11/23/2021"
output: 
  html_document: 
    theme: cerulean
---

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

# attach packages
library(tidyverse)
library(janitor)
library(broom)
library(here)
```

## Read in the data

```{r}
lizards <- read_csv(here("data", "lizards.csv"))
```

Find the counts by species, what are most common lizards

```{r}
lizards %>% 
  count(spp) %>% 
  arrange(-n)  # this will change arrangement to list high to low
```

Found most common was UTST (side-blotched)

## Make a subset of the data

Question relating to tail condition 

- UTST lizards
- sites: IBPE (enclosure grassland), RABB (rabbit mesquite), "EAST" (east tarbush)
- tail condition is recorded (W = whole, B = broken)

```{r}
utst <- lizards %>%
  filter(spp == "UTST") %>% 
  filter(site %in% c("IBPE", "RABB", "EAST")) %>% 
  filter(tail %in% c("B", "W"))
```

remember can use `unique` with $ to see what is actually left

## Find counts and proportions of lizards with broken or whole tails at each site (side blotched only)

```{r}
utst %>% 
  count(site, tail)  # not a super pretty table, lets fix it to be tidy

# could also use pivot wider
```

use the `janitor` package with the `tabyl()` function and the `adorn_percentages()` function
easier to compare across groups than long table formats

```{r}
tail_counts <- utst %>% 
  janitor::tabyl(site, tail)

tail_proportions <- tail_counts %>% 
  adorn_percentages() %>%  # note default is to calculate proportions by rows 
  adorn_pct_formatting(digits = 2) %>%  # and update digits
  adorn_ns() # and add sample size

# but nice to have one table with both - nice way to compare proportions and sample size
# could update with kableExtra to be super pretty 

```


### is there a significant association in broken tails by sites?

first need to clean up the table so that it converts the first column into row names 
this will give us a contingency table with just the numbers ready to analyze

## Convert to contingency table

Use `column to rowname()`
then we're left with just contingency values, which is what we want to analye

```{r}
lizard_ct <- tail_counts %>% 
  column_to_rownames(var = "site")  # different than normal syntax for variable

```

## run a chi-square

Q: are tail breakage and site independent?  
is there a significant association between these variables?
would assume tail breakage proportion to be same across site; null hypothesis of independence 

```{r}
lizard_x2 <- chisq.test(lizard_ct)
lizard_x2

lizard_tidy <- tidy(lizard_x2) # this makes it easier to refer in line referencing!
```

degrees of freedom = number or rows-1 * number of columns-1
yes, we have enough information to reject the null hypothesis
significant association between tail breakage and site
site and lizard tail breakage are not independent

There is a significant association (i.e., non-independence) between site and lizard tail breakage ($\chi$^2^(`r lizard_tidy$parameter`) = `r round(lizard_tidy$statistic,2)`, *p* = `r round(lizard_tidy$p.value, 3)`).

